---
title: "Multidimensional Correlates of Protection with seroCOP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multidimensional Correlates of Protection with seroCOP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8, fig.height = 6)
```

# Joint Biomarker CoP

`SeroMulti` fits one joint model to 2--4 antibody biomarkers measured for the
same person and time point. Each biomarker has its own EC50 and positive slope.
For $D$ biomarkers, the joint signal is

$$
\eta = \sum_{d=1}^{D} \text{slope}_d (\text{titre}_d - \text{ec50}_d).
$$

The CoP rises when the combined biomarker signal increases. This vignette uses
two biomarkers, so the fitted CoP can be inspected as a surface. Three- and
four-biomarker models use the same interface, but are evaluated at chosen
biomarker combinations rather than drawn as a surface.

```{r setup}
library(seroCOP)
library(ggplot2)
set.seed(2026)
```

# Simulate a Two-Biomarker Study

```{r simulate-data}
n <- 300
titres <- cbind(
  IgG = rnorm(n, mean = 2.2, sd = 1.0),
  Neutralisation = rnorm(n, mean = 1.8, sd = 0.9)
)

true_floor <- 0.05
true_ceiling <- 0.75
true_ec50 <- c(IgG = 2.0, Neutralisation = 1.5)
true_slope <- c(IgG = 1.1, Neutralisation = 0.9)
eta <- as.vector((titres - rep(true_ec50, each = n)) %*% true_slope)
infection_probability <- true_ceiling *
  (plogis(-eta) * (1 - true_floor) + true_floor)
infected <- rbinom(n, size = 1, prob = infection_probability)

head(data.frame(titres, infected))
```

# Visualise the Joint CoP Surface

The simulation gives a known surface before fitting. It provides a visual
reference for the fitted posterior surface below.

```{r true-surface}
surface_grid <- expand.grid(
  IgG = seq(min(titres[, "IgG"]), max(titres[, "IgG"]), length.out = 80),
  Neutralisation = seq(min(titres[, "Neutralisation"]),
                       max(titres[, "Neutralisation"]), length.out = 80)
)
grid_eta <- as.vector((as.matrix(surface_grid) -
  rep(true_ec50, each = nrow(surface_grid))) %*% true_slope)
grid_risk <- true_ceiling * (plogis(-grid_eta) * (1 - true_floor) + true_floor)
surface_grid$cop <- 1 - grid_risk / true_ceiling

ggplot(surface_grid, aes(x = IgG, y = Neutralisation, fill = cop)) +
  geom_raster(interpolate = TRUE) +
  geom_contour(aes(z = cop), color = "white") +
  scale_fill_viridis_c(name = "CoP") +
  labs(title = "Simulated joint correlate-of-protection surface") +
  theme_minimal()
```

# Fit SeroMulti

```{r fit-model, eval=FALSE}
model <- SeroMulti$new(titre = titres, infected = infected)
model$fit_model(chains = 4, iter = 2000, warmup = 1000, cores = 4)
```

# Plot the Fitted Surface

After fitting, `plot_surface()` summarises posterior CoP draws across a grid.
It returns a ggplot filled-contour plot by default; `type = "persp"` draws a
base-R perspective surface.

```{r fitted-surface, eval=FALSE}
model$plot_surface(type = "contour")
model$plot_surface(type = "persp")
```

# Three and Four Biomarkers

For higher-dimensional models, supply a 3- or 4-column matrix. Posterior CoP
values can then be obtained for observed or selected biomarker combinations.

```{r higher-dimensions, eval=FALSE}
three_marker_titres <- cbind(titres, IgA = rnorm(n, 1.6, 0.8))
three_marker_model <- SeroMulti$new(three_marker_titres, infected)
three_marker_model$fit_model(chains = 4, iter = 2000, warmup = 1000, cores = 4)

posterior_cop <- three_marker_model$predict_protection(three_marker_titres)
colMeans(posterior_cop)
```