---
title: "SeroCOP Overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SeroCOP Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 8,
  fig.height = 5
)
```

# Overview of SeroCOP

The `seroCOP` package provides Bayesian tools for analysing **correlates of
protection (CoP)** ŌĆö the relationship between an individual's antibody level
and their risk of infection. Models are fit using
[brms](https://paul-buerkner.github.io/brms/) (backed by Stan) and expose
posterior predictions, performance metrics, and publication-quality plots
through two R6 classes:

| Class | Purpose |
|-------|---------|
| `SeroCOP` | Single biomarker; optionally hierarchical by group |
| `SeroCOPMulti` | Multiple biomarkers fitted and compared simultaneously |

---

## Model Description

The probability of infection is modelled as a four-parameter logistic curve:

$$
P(\text{infection} \mid t) =
  \text{ceiling} \times
  \left[\text{inv\_logit}\!\bigl(-\text{slope}\,(t - \text{ec50})\bigr)
    \times (1 - \text{floor}) + \text{floor}\right]
$$

| Parameter | Interpretation |
|-----------|---------------|
| `floor`   | Residual infection risk at the highest titres (near-complete protection) |
| `ceiling` | Maximum infection risk at the lowest titres (baseline attack rate) |
| `ec50`    | Titre at the inflection point of the curve |
| `slope`   | Steepness of the doseŌĆōresponse (higher = sharper transition) |

---

# Quick Start

```{r quick-start, eval=FALSE}
library(seroCOP)

# Simulate data
set.seed(123)
n        <- 300
titre    <- rnorm(n, mean = 2, sd = 1.5)
prob     <- 0.05 + 0.65 / (1 + exp(2 * (titre - 1.5)))
infected <- rbinom(n, 1, prob)

# Initialise the model
model <- SeroCOP$new(titre = titre, infected = infected)

# Fit (brms / Stan backend)
model$fit_model(chains = 4, iter = 2000, warmup = 1000, cores = 4)

# Inspect parameters
model$summary()

# Performance metrics
model$get_metrics()

# Plots
model$plot_curve()
model$plot_roc()
model$plot_posteriors()
```

---

# Prior Distributions

Default priors are weakly informative and data-driven:

| Parameter | Default prior | Rationale |
|-----------|--------------|-----------|
| `floor`   | Beta(1, 9)   | Favours low residual risk |
| `ceiling` | Beta(9, 1)   | Favours high baseline risk |
| `ec50`    | Normal(mid, range/4) | Centred on the data range |
| `slope`   | Normal(0, 2) truncated Ōēź 0 | Weakly informative |

Override priors with `definePrior()`:

```{r define-prior, eval=FALSE}
model$definePrior(
  floor_alpha   = 2,   floor_beta    = 18,   # E[floor]   Ōēł 0.10
  ceiling_alpha = 18,  ceiling_beta  = 2,    # E[ceiling] Ōēł 0.90
  ec50_mean     = 2.0, ec50_sd       = 1.0,  # Informative ec50 prior
  slope_mean    = 0,   slope_sd      = 1     # Tighter slope prior
)
```

`definePrior()` can be called any time **before** `fit_model()` and prints a
summary of the updated priors.

---

# Observation Weights

Pass a non-negative numeric vector to `SeroCOP$new(weights = ...)` to
up-/down-weight individual observations in the likelihood:

```{r weights-example, eval=FALSE}
# Case-control design: up-weight confirmed cases
w <- ifelse(infected == 1, 1.0, 0.5)

w_model <- SeroCOP$new(titre = titre, infected = infected, weights = w)
w_model$fit_model(chains = 4, iter = 2000, warmup = 1000)
```

---

# Hierarchical Models (Multiple Groups)

Add a `group` variable to estimate group-specific `ec50` and `slope` while
sharing a common `floor` (and optionally `ceiling`):

```{r hier-example, eval=FALSE}
age_group <- sample(c("Young", "Middle", "Old"), n, replace = TRUE)

hier <- SeroCOP$new(titre = titre, infected = infected, group = age_group)

# ceiling_group = FALSE (default): ceiling shared across groups
# ceiling_group = TRUE:  ceiling also allowed to vary by group
hier$fit_model(chains = 4, iter = 2000, warmup = 1000,
               ceiling_group = FALSE)

# Group-specific parameter table
hier$extract_group_parameters()

# Fitted risk curves with marginalised overlay
hier$plot_group_curves(
  weights_pop = c(Young = 0.45, Middle = 0.35, Old = 0.20)
)

# Correlates of protection curves
hier$plot_cop_curves(
  weights_pop = c(Young = 0.45, Middle = 0.35, Old = 0.20)
)

# Per-group LOO is computed automatically; accessed via:
hier$loo_by_group
```

---

# Multi-Biomarker Analysis

`SeroCOPMulti` fits one independent `SeroCOP` model per biomarker and
provides comparison tools:

```{r multi-example, eval=FALSE}
# titre is a matrix: rows = observations, cols = biomarkers
titre_mat <- cbind(IgG = titre_IgG, IgA = titre_IgA, Neutralisation = titre_Neut)

multi <- SeroCOPMulti$new(
  titre           = titre_mat,
  infected        = infected,
  biomarker_names = c("IgG", "IgA", "Neutralisation")
)

multi$fit_all(chains = 4, iter = 2000, warmup = 1000, cores = 4)

# Compare biomarkers (AUC, Brier, LOO-ELPD)
multi$compare_biomarkers()
multi$plot_comparison()

# All fitted risk curves side-by-side
multi$plot_all_curves()
```

---

# Performance Metrics

`get_metrics()` returns and prints:

| Metric | Description |
|--------|-------------|
| `roc_auc` | Area under the ROC curve (marginalised predictions) |
| `roc_auc_within_group` | Weighted within-group AUC (hierarchical only) |
| `brier_score` | Mean squared error of predicted probabilities |
| `loo_elpd` / `loo_se` | LOO expected log predictive density ┬▒ SE |
| `loo_by_group` | Per-group LOO ELPD (hierarchical only) |

---

# Extracting Parameters

`extract_parameters()` wraps `brms::as_draws_df()` and returns a tidy
data frame of posterior mean, median, and credible intervals:

```{r extract-params-example, eval=FALSE}
params <- extract_parameters(model, prob = 0.95)
print(params)
#   parameter  mean median  lower  upper
# 1     floor 0.052  0.050  0.010  0.110
# 2   ceiling 0.681  0.683  0.612  0.748
# 3      ec50 1.504  1.503  1.363  1.641
# 4     slope 1.983  1.978  1.701  2.275
```

---

# Predicting Protection

The **correlate of protection** at titre $t$ is:

$$\text{COP}(t) = 1 - \frac{P(\text{infection} \mid t)}{\text{ceiling}}$$

```{r cop-example, eval=FALSE}
# Matrix of posterior samples (rows = draws, cols = observations)
cop_samples <- model$predict_protection(newdata = seq(0, 5, 0.1))

cop_mean  <- colMeans(cop_samples)
cop_lower <- apply(cop_samples, 2, quantile, 0.025)
cop_upper <- apply(cop_samples, 2, quantile, 0.975)
```

---

# Session Info

```{r session-info}
sessionInfo()
```