---
title: "Weighted Multi-Biomarker Analysis with seroCOP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Weighted Multi-Biomarker Analysis with seroCOP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)
```

# Introduction

This vignette demonstrates how to supply biomarker-specific **observation
weights** to `SeroCOPMulti`. Weights scale each observation's contribution to
the Bernoulli log-likelihood in Stan — effectively giving more or less influence
to individual observations during model fitting.

A common use case is **case-control up-weighting**: infected cases are given
full weight (`1`), while non-infected controls are given reduced, potentially
varying weights (`0`–`0.8`) to reflect sub-sampling, survey design, or
uncertainty in the non-infected group.

We demonstrate this with three simulated biomarkers:

| Biomarker | Correlate of protection? |
|-----------|--------------------------|
| IgG       | Strong                   |
| IgA       | Moderate                 |
| Non-specific | None                  |

## Load the package

```{r setup}
library(seroCOP)
library(ggplot2)
set.seed(2025)
```

# Simulate Multi-Biomarker Data

```{r simulate-data}
n <- 250

# --- Biomarker 1: IgG (strong CoP) ---
titre_IgG <- rnorm(n, mean = 2.5, sd = 1.2)
prob_IgG  <- 0.02 + 0.68 / (1 + exp(2.5 * (titre_IgG - 2.0)))

# --- Biomarker 2: IgA (moderate CoP) ---
titre_IgA <- rnorm(n, mean = 1.8, sd = 1.5)
prob_IgA  <- 0.15 + 0.55 / (1 + exp(1.0 * (titre_IgA - 1.5)))

# --- Biomarker 3: Non-specific (no CoP) ---
titre_Nonspec <- rnorm(n, mean = 3.0, sd = 1.0)
prob_Nonspec  <- rep(0.35, n)

# Infection outcome (driven primarily by IgG)
prob_combined <- 0.5 * prob_IgG + 0.3 * prob_IgA + 0.2 * prob_Nonspec
infected      <- rbinom(n, 1, prob_combined)

cat(sprintf("Infection rate: %.1f%% (%d / %d)\n",
            mean(infected) * 100, sum(infected), n))

# Combine titres into a matrix
titres <- cbind(IgG = titre_IgG, IgA = titre_IgA, NonSpecific = titre_Nonspec)
```

# Construct Observation Weights

Infected observations receive a weight of **1** (full influence). Non-infected
observations receive weights drawn uniformly from **0 to 0.8**, simulating a
scenario where controls are sub-sampled or partially trusted. Weights are
biomarker-specific: each column of the weight matrix corresponds to one column
of the `titre` matrix.

```{r build-weights}
# Initialise weight matrix — same dimensions as titres
weights_mat <- matrix(1.0, nrow = n, ncol = ncol(titres))
colnames(weights_mat) <- colnames(titres)

# Non-infected rows get uniform random weights in [0, 0.8]
non_infected_idx <- which(infected == 0)

for (j in seq_len(ncol(weights_mat))) {
  weights_mat[non_infected_idx, j] <- runif(length(non_infected_idx),
                                             min = 0, max = 0.8)
}

# Quick summary
cat("Weight summary by biomarker:\n")
print(round(apply(weights_mat, 2, summary), 3))
```

```{r weight-dist-plot, fig.cap="Distribution of observation weights split by infection status"}
weight_df <- data.frame(
  weight   = as.vector(weights_mat),
  biomarker = rep(colnames(weights_mat), each = n),
  status   = rep(ifelse(infected == 1, "Infected", "Not infected"), ncol(weights_mat))
)

ggplot(weight_df, aes(x = weight, fill = status)) +
  geom_histogram(binwidth = 0.05, position = "identity", alpha = 0.6) +
  facet_wrap(~biomarker) +
  scale_fill_manual(values = c("Infected" = "#E74C3C", "Not infected" = "#3498DB")) +
  labs(
    title = "Observation weight distributions by infection status and biomarker",
    x = "Weight", y = "Count", fill = "Status"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

# Fit the Weighted Multi-Biomarker Model

Pass the weight matrix to `SeroCOPMulti$new()`. Each column is automatically
routed to the corresponding biomarker's `SeroCOP` sub-model, where it enters
Stan as `infected | weights(obs_weights)`, scaling each observation's
contribution to the log-posterior.

```{r fit-weighted, eval = FALSE}
weighted_model <- SeroCOPMulti$new(
  titre    = titres,
  infected = infected,
  weights  = weights_mat
)

weighted_model$fit_all(chains = 4, iter = 2000, warmup = 1000)
```

```{r fit-weighted-run, include = FALSE, eval = FALSE}
# (Separated so the output above renders cleanly in the vignette)
weighted_model$fit_all(chains = 4, iter = 2000, warmup = 1000)
```

# Compare Weighted vs. Unweighted Models

Fitting an unweighted model on the same data lets us see how down-weighting
non-infected controls changes parameter estimates and performance metrics.

```{r fit-unweighted, eval = FALSE}
unweighted_model <- SeroCOPMulti$new(
  titre    = titres,
  infected = infected
)

unweighted_model$fit_all(chains = 4, iter = 2000, warmup = 1000)
```

## Performance metrics

```{r compare-metrics, eval = FALSE}
cat("=== Weighted model ===\n")
weighted_metrics <- weighted_model$compare_biomarkers()

cat("\n=== Unweighted model ===\n")
unweighted_metrics <- unweighted_model$compare_biomarkers()
```

## Fitted curves — weighted model

```{r plot-weighted-curves, eval = FALSE}
weighted_model$plot_all_curves() +
  labs(title = "Weighted: Correlates of Risk Curves by Biomarker")
```

## Fitted curves — unweighted model

```{r plot-unweighted-curves, eval = FALSE}
unweighted_model$plot_all_curves() +
  labs(title = "Unweighted: Correlates of Risk Curves by Biomarker")
```

# Interpreting the Results

Downweighting non-infected observations has several predictable effects:

- **Steeper curves**: the model leans more heavily on the signal from cases,
  which can sharpen the estimated dose-response.
- **Wider credible intervals**: fewer effective observations for non-infected
  controls means greater posterior uncertainty in the floor and ec50 parameters.
- **Higher AUC (potentially)**: if non-infected controls are noisy or
  heterogeneous, reducing their influence can improve discrimination.

The `IgG` biomarker is expected to show the largest change because it has the
clearest dose-response, so the relative up-weighting of cases amplifies that
signal most noticeably.

# Summary

```{r summary-table, eval = FALSE}
summary_df <- merge(
  weighted_metrics[, c("biomarker", "auc", "brier", "loo_elpd")],
  unweighted_metrics[, c("biomarker", "auc", "brier", "loo_elpd")],
  by = "biomarker", suffixes = c("_weighted", "_unweighted")
)

knitr::kable(summary_df, digits = 3,
             caption = "Weighted vs. Unweighted model performance by biomarker")
```

Key takeaways:

- Weights are a **matrix** matching `titre` dimensions — every biomarker can
  have a distinct weighting scheme.
- A plain **vector** is also accepted and broadcast across all biomarkers.
- Infected observations are given weight `1`; non-infected controls receive
  random weights in `[0, 0.8]`.
- The mechanics are implemented in Stan via
  `target += w[i] * bernoulli_lpmf(y[i] | p[i])`.
- When `weights = NULL` (the default), all observations receive equal weight
  and behaviour is identical to the standard `SeroCOPMulti` workflow.
